---
title: "Variance Decomposition (PEcAn Sobol)"
date: today
format:
  html:
    self-contained: true
    toc: true
    toc-depth: 2
    code-fold: true
    theme: cosmo
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false

library(config)
library(dplyr)
library(ggplot2)
library(here)
library(knitr)
library(readr)
library(scales)

here::i_am("analysis/variance_decomposition_pecan_sobol.qmd")
cfg <- config::get(file = here::here("000-config.yml"))
data_dir <- here::here(cfg$paths$data_dir, "pecan_sobol")

site_partition <- readr::read_csv(
  file.path(data_dir, "variance_partition_site_level.csv"),
  show_col_types = FALSE
)

param_partition <- readr::read_csv(
  file.path(data_dir, "variance_partition_parameters.csv"),
  show_col_types = FALSE
)

type_summary <- readr::read_csv(
  file.path(data_dir, "variance_partition_summary.csv"),
  show_col_types = FALSE
)

param_summary <- readr::read_csv(
  file.path(data_dir, "variance_partition_params_summary.csv"),
  show_col_types = FALSE
)

pretty_source <- function(x) {
  dplyr::case_when(
    x == "param" ~ "Parameters",
    x == "met" ~ "Meteorology",
    x == "poolinitcond" ~ "Initial Conditions",
    x == "events" ~ "Events / Management",
    x == "interaction" ~ "Interactions",
    TRUE ~ x
  )
}

ordered_source_levels <- function(observed, preferred) {
  observed <- unique(observed[!is.na(observed) & nzchar(observed)])
  c(
    preferred[preferred %in% observed],
    sort(setdiff(observed, preferred))
  )
}

site_partition <- site_partition |>
  dplyr::mutate(
    source = pretty_source(.data$category),
    site_label = dplyr::if_else(
      !is.na(.data$site_id) & nzchar(.data$site_id),
      .data$site_id,
      .data$runid
    )
  )

type_summary <- type_summary |>
  dplyr::mutate(source = pretty_source(.data$category))

site_source_levels <- ordered_source_levels(
  site_partition$source,
  c(
    "Interactions",
    "Events / Management",
    "Initial Conditions",
    "Meteorology",
    "Parameters"
  )
)

summary_source_levels <- ordered_source_levels(
  type_summary$source,
  c(
    "Parameters",
    "Meteorology",
    "Initial Conditions",
    "Events / Management"
  )
)

theme_set(
  theme_minimal(base_size = 12) +
    theme(
      panel.grid.minor = element_blank(),
      legend.position = "bottom"
    )
)
```

```{r}
#| label: findings
#| include: false

dominant_types <- type_summary |>
  dplyr::group_by(.data$variable) |>
  dplyr::slice_max(order_by = .data$mean_frac, n = 1, with_ties = FALSE) |>
  dplyr::ungroup()

interaction_mean <- site_partition |>
  dplyr::filter(.data$category == "interaction") |>
  dplyr::summarize(mean_frac = mean(.data$frac_of_total, na.rm = TRUE)) |>
  dplyr::pull(.data$mean_frac)

top_params <- param_summary |>
  dplyr::group_by(.data$response_var) |>
  dplyr::slice_max(order_by = .data$mean_frac, n = 5, with_ties = FALSE) |>
  dplyr::ungroup()
```

# Overview

This report consumes PEcAn-generated `sobol.indices.*.Rdata` outputs and keeps the existing local sensitivity analysis only as a within-`param` drilldown. Type-level variance decomposition is therefore taken directly from the PEcAn Sobol workflow, not recomputed here.

`uncertainty` reads factor/type metadata from the PEcAn-saved Sobol objects and uses the canonical PEcAn filenames only for file-level keys such as run id, variable, and year bounds.

- Mean interaction share across all site-variable combinations: `r percent(interaction_mean, accuracy = 0.1)`.
- Dominant type by variable:

```{r}
knitr::kable(
  dominant_types |>
    dplyr::transmute(
      variable,
      dominant_source = .data$source,
      mean_fraction = scales::percent(.data$mean_frac, accuracy = 0.1)
    ),
  align = c("l", "l", "r")
)
```

# Type-Level Sobol Results

The figure below shows the PEcAn type-level decomposition by site. Categories are the canonical PEcAn factor ids rendered into readable labels only for the report.

```{r}
ggplot(
  site_partition |>
    dplyr::mutate(
      source = factor(.data$source, levels = site_source_levels)
    ),
  aes(x = .data$site_label, y = .data$frac_of_total, fill = .data$source)
) +
  geom_col(position = "fill", width = 0.85) +
  facet_wrap(~variable, ncol = 1, scales = "free_x") +
  scale_y_continuous(labels = scales::percent) +
  scale_fill_brewer(palette = "Set2", name = "Source") +
  labs(
    x = "Site / Run",
    y = "Fraction of total variance"
  ) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```

```{r}
ggplot(
  type_summary |>
    dplyr::mutate(
      source = factor(.data$source, levels = summary_source_levels)
    ),
  aes(x = .data$source, y = .data$mean_frac, fill = .data$source)
) +
  geom_col() +
  geom_errorbar(
    aes(
      ymin = pmax(.data$mean_frac - .data$sd_frac, 0),
      ymax = .data$mean_frac + .data$sd_frac
    ),
    width = 0.2
  ) +
  facet_wrap(~variable, scales = "free_y") +
  scale_y_continuous(labels = scales::percent) +
  scale_fill_brewer(palette = "Set2", guide = "none") +
  labs(
    x = NULL,
    y = "Mean fraction of total variance"
  ) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```

# Parameter Drilldown Within `param`

These results keep the existing local OAT attribution logic, but only after the PEcAn Sobol workflow has determined the size of the overall `param` contribution.

```{r}
if (nrow(param_summary) == 0) {
  cat("No parameter-level decomposition was produced.\n")
} else {
  ggplot(
    top_params,
    aes(
      x = reorder(.data$parameter, .data$mean_frac),
      y = .data$mean_frac,
      fill = .data$response_var
    )
  ) +
    geom_col(show.legend = FALSE) +
    coord_flip() +
    facet_wrap(~response_var, scales = "free_y") +
    scale_y_continuous(labels = scales::percent) +
    labs(
      x = "Parameter",
      y = "Mean fraction of total variance"
    )
}
```

```{r}
if (nrow(top_params) > 0) {
  knitr::kable(
    top_params |>
      dplyr::transmute(
        response_var,
        parameter,
        mean_fraction = scales::percent(.data$mean_frac, accuracy = 0.1)
      ),
    align = c("l", "l", "r")
  )
}
```

# Data Products

This report uses the dedicated PEcAn-backed outputs written under `data/pecan_sobol/`:

- `type_level_indices.csv`
- `ensemble_variance.csv`
- `variance_partition_site_level.csv`
- `variance_partition_parameters.csv`
- `variance_partition_summary.csv`
- `variance_partition_params_summary.csv`
